---
title: "Liu (2009) Simulation Study"
author: "Shozen Dan"
date: "2025-11-23"
output: html_document
---

```{r setup}
library(dplyr)
library(ggplot2)
library(cmdstanr)
library(bayesplot)
library(mvtnorm)
library(purrr)

REPO_DIR <- "~/Imperial/0_Research/modular_bayes/"
```

```{r experiment-config}
set.seed(1234)

config <- list(
  n1 = 25,
  n2 = 50,
  mu = 0,
  theta = 1,
  s1 = 1,
  s2 = 1
)
```

```{r}
# Function to generate data for the simulation study
generate_data <- function(n1, n2, mu, theta, s1, s2) {
  y1 <- rnorm(n1, mean = mu, sd = s1)
  y2 <- rnorm(n2, mean = mu + theta, sd = s2)

  return(list(y1 = y1, y2 = y2))
}

set.seed(1234)
data <- generate_data(
  n1 = config$n1,
  n2 = config$n2,
  mu = config$mu,
  theta = config$theta,
  s1 = config$s1,
  s2 = config$s2
)
```

```{r sanity-check}
ggplot() +
  geom_histogram(aes(x = data$y1), bins = 10, fill = "blue", alpha = 0.5) +
  geom_histogram(aes(x = data$y2), bins = 10, fill = "red", alpha = 0.5) +
  labs(title = "Histograms of Simulated Data", x = "Value", y = "Frequency") +
  theme_bw()
```

## Fit a model on the first dataset
```{r message=FALSE}
stan_data <- list(
  n1 = config$n1,
  y1 = data$y1
)

model_m1 <- cmdstan_model("../../stan/liu_m1.stan")
fit_m1 <- model_m1$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1234,
  show_messages = FALSE
)
```

Summary of the posterior on the first model.
```{r}
summary_m1 <- fit_m1$summary("mu")
mu_m1_mean <- summary_m1$mean
mu_m1_median <- summary_m1$median

print(summary_m1)
```

## Fit a full model in Stan
```{r prepare-data-for-stan}
stan_data <- list(
  n1 = config$n1,
  n2 = config$n2,
  y1 = data$y1,
  y2 = data$y2
)
```

```{r fit-model, message=FALSE}
model_full <- cmdstan_model("../../stan/liu_full.stan")
fit_full <- model_full$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 500,
  iter_sampling = 500,
  seed = 1234,
  show_messages = FALSE
)
```

Print the posterior summaries for $\mu$ and $\theta$.
```{r posterior-summary}
post_full <- fit_full$draws()
summary_full <- fit_full$summary(c("mu", "theta"))

print(summary_full)
```

### Importance sampling step
1. Compute the weights for each sample in the MCMC
2. Self-normalise the weights
3. Multi-nomial resampling

```{r}
# Extract posterior draws as numeric vectors
mu_draws <- as.numeric(post_full[, , "mu"])
theta_draws <- as.numeric(post_full[, , "theta"])

# Create particle matrix for cleaner interface
particles <- cbind(
  mu = mu_draws,
  tilde_theta = theta_draws, # Create an auxiliary parameter
  theta = theta_draws
)
```

```{r}
source("../../R/liu/compute_log_weights.R")
source("../../R/liu/is_resample.R")
source("../../R/liu/log_posterior.R")
source("../../R/liu/mh_step.R")
source("../../R/liu/rejuvenate.R")
source("../../R/liu/smc_step.R")
source("../../R/liu/run_smc.R")
```

## Run Full SMC Algorithm
### Execute SMC over the complete tempering schedule
```{r run-full-smc}
# Run complete SMC algorithm
smc_full_result <- run_smc(
  particles_init = particles,
  y1 = data$y1,
  y2 = data$y2,
  s1 = config$s1,
  s2 = config$s2,
  eta_schedule = c(1,0.5,0.3,0.2,0.1,0.05,0.01,0.001,0.0001,1e-5,0),
  n_mh_steps = 100,
  proposal_sd = 1,
  verbose = TRUE
)
```

### Compare with full Stan posterior
```{r compare-results}
# Extract final SMC particles
final_particles <- smc_full_result$final_particles

# Compare with Stan results
cat("=== Comparison: SMC vs Stan Full Model ===\n\n")

cat("Stan (full data from start):\n")
cat(sprintf(
  "  Mu:    mean=%.4f, sd=%.4f\n",
  mean(mu_draws), sd(mu_draws)
))
cat(sprintf(
  "  Theta: mean=%.4f, sd=%.4f\n\n",
  mean(theta_draws), sd(theta_draws)
))

cat("SMC (sequential tempering):\n")
cat(sprintf(
  "  Mu:    mean=%.4f, sd=%.4f\n",
  mean(final_particles[, "mu"]), sd(final_particles[, "mu"])
))
cat(sprintf(
  "  Theta: mean=%.4f, sd=%.4f\n\n",
  mean(final_particles[, "theta"]), sd(final_particles[, "theta"])
))

# Print summary statistics table
cat("=== Summary Statistics Across Tempering Steps ===\n\n")
print(smc_full_result$summary_stats, digits = 4)
```

### Visualize SMC evolution
Theoretically ideal mu and theta
```{r}
mu_ideal <- mean(data$y1)
mu_sd_ideal <- var(data$y1) / config$n1

rho <- var(data$y2) * 0.5**2 /(var(data$y2)/config$n2 + 0.5**2)
theta_ideal <- rho*(mean(data$y2) - mu_ideal)
theta_sd_ideal <- rho / config$n2
```

```{r visualize-smc-evolution, fig.width=12, fig.height=10}
# Plot evolution of posterior estimates
par(mfrow = c(2, 3))

# Plot 1: Mu mean trajectory
plot(smc_full_result$summary_stats$step,
  smc_full_result$summary_stats$mu_mean,
  type = "b", pch = 16, col = "blue",
  xlab = "SMC Step", ylab = "Mean of Mu",
  main = "Evolution of Mu Mean",
  ylim = c(-.3, 0.01)
)
abline(h = mean(mu_draws), col = "red", lty = 2, lwd = 2)
abline(h = config$mu, col = "green", lty = 2, lwd = 2)
abline(h = mu_ideal, col = "orange", lty = 2, lwd = 2)
legend("topright",
  legend = c("SMC", "Stan", "True", "Ideal"),
  col = c("blue", "red", "green", "orange"),
  lty = c(1, 2, 2, 2), pch = c(16, NA, NA, NA)
)

# Plot 2: Theta mean trajectory
plot(smc_full_result$summary_stats$step,
  smc_full_result$summary_stats$theta_mean,
  type = "b", pch = 16, col = "blue",
  xlab = "SMC Step", ylab = "Mean of Theta",
  main = "Evolution of Theta Mean",
  ylim = c(0, 1.2)
)
abline(h = mean(theta_draws), col = "red", lty = 2, lwd = 2)
abline(h = config$theta, col = "green", lty = 2, lwd = 2)
abline(h = theta_ideal, col = "orange", lty = 2, lwd = 2)
legend("topright",
  legend = c("SMC", "Stan", "True", "Ideal"),
  col = c("blue", "red", "green", "orange"),
  lty = c(1, 2, 2), pch = c(16, NA, NA)
)

# Plot 3: Eta schedule
plot(smc_full_result$summary_stats$step,
  smc_full_result$summary_stats$eta,
  type = "b", pch = 16, col = "darkgreen",
  xlab = "SMC Step", ylab = "Eta (Tempering Parameter)",
  main = "Tempering Schedule"
)

# Plot 4: Mu SD trajectory
plot(smc_full_result$summary_stats$step,
  smc_full_result$summary_stats$mu_sd,
  type = "b", pch = 16, col = "blue",
  xlab = "SMC Step", ylab = "SD of Mu",
  main = "Evolution of Mu SD",
  ylim = c(0, 0.4)
)
abline(h = sd(mu_draws), col = "red", lty = 2, lwd = 2)
abline(h = mu_sd_ideal, col = "orange", lty = 2, lwd = 2)
legend("topright",
  legend = c("SMC", "Stan", "Ideal"),
  col = c("blue", "red", "orange"),
  lty = c(1, 2, 2), pch = c(16, NA, NA)
)

# Plot 5: Theta SD trajectory
plot(smc_full_result$summary_stats$step,
  smc_full_result$summary_stats$theta_sd,
  type = "b", pch = 16, col = "blue",
  xlab = "SMC Step", ylab = "SD of Theta",
  main = "Evolution of Theta SD",
  ylim = c(0, 0.4)
)
abline(h = sd(theta_draws), col = "red", lty = 2, lwd = 2)
abline(h = theta_sd_ideal, col = "orange", lty = 2, lwd = 2)
legend("topright",
  legend = c("SMC", "Stan", "Ideal"),
  col = c("blue", "red", "orange"),
  lty = c(1, 2, 2), pch = c(16, NA, NA)
)

# Plot 6: Combined trajectory in parameter space
plot(smc_full_result$summary_stats$mu_mean,
  smc_full_result$summary_stats$theta_mean,
  type = "b", pch = 16, col = "blue",
  xlab = "Mean of Mu", ylab = "Mean of Theta",
  main = "SMC Trajectory in Parameter Space"
)
points(mean(mu_draws), mean(theta_draws), col = "red", pch = 17, cex = 2)
points(config$mu, config$theta, col = "green", pch = 18, cex = 2)
text(smc_full_result$summary_stats$mu_mean,
  smc_full_result$summary_stats$theta_mean,
  labels = 0:10, pos = 3, cex = 0.7, col = "blue"
)
legend("topright",
  legend = c("SMC path", "Stan", "True"),
  col = c("blue", "red", "green"),
  pch = c(16, 17, 18)
)

par(mfrow = c(1, 1))
```

### Compare distributions at key steps
```{r compare-distributions, fig.width=12, fig.height=8}
# Plot histograms at eta = 1, 0.5, and 0
par(mfrow = c(2, 3))

# Eta = 1 (step 0)
step_idx <- 1
hist(smc_full_result$particle_history[[step_idx]][, "mu"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Mu at eta=%.1f", smc_full_result$eta_schedule[step_idx]),
  xlab = "mu", freq = FALSE
)
lines(density(mu_draws), col = "red", lwd = 2)

hist(smc_full_result$particle_history[[step_idx]][, "theta"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Theta at eta=%.1f", smc_full_result$eta_schedule[step_idx]),
  xlab = "theta", freq = FALSE
)
lines(density(theta_draws), col = "red", lwd = 2)

# Eta = 0.5 (step 5)
step_idx <- 6
hist(smc_full_result$particle_history[[step_idx]][, "mu"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Mu at eta=%.1f", smc_full_result$eta_schedule[step_idx]),
  xlab = "mu", freq = FALSE
)
lines(density(mu_draws), col = "red", lwd = 2)

hist(smc_full_result$particle_history[[step_idx]][, "theta"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Theta at eta=%.1f", smc_full_result$eta_schedule[step_idx]),
  xlab = "theta", freq = FALSE
)
lines(density(theta_draws), col = "red", lwd = 2)

# Eta = 0 (final step)
step_idx <- 11
hist(smc_full_result$particle_history[[step_idx]][, "mu"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Mu at eta=%.1f (final)", smc_full_result$eta_schedule[step_idx]),
  xlab = "mu", freq = FALSE
)
lines(density(mu_draws), col = "red", lwd = 2)
legend("topright",
  legend = c("SMC", "Stan"),
  fill = c(rgb(0, 0, 1, 0.5), NA),
  border = c("black", NA),
  lty = c(NA, 1), col = c(NA, "red"), lwd = c(NA, 2)
)

hist(smc_full_result$particle_history[[step_idx]][, "theta"],
  breaks = 30, col = rgb(0, 0, 1, 0.5),
  main = sprintf("Theta at eta=%.1f (final)", smc_full_result$eta_schedule[step_idx]),
  xlab = "theta", freq = FALSE
)
lines(density(theta_draws), col = "red", lwd = 2)
legend("topright",
  legend = c("SMC", "Stan"),
  fill = c(rgb(0, 0, 1, 0.5), NA),
  border = c("black", NA),
  lty = c(NA, 1), col = c(NA, "red"), lwd = c(NA, 2)
)

par(mfrow = c(1, 1))
```