---
title: "vanilla_gaussian_example"
output: html_document
---

## Libraries
```{r message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(loo)
library(iwmm)

set_theme(theme_bw())
```

## Setup
```{r}
# True value of phi
PHI <- 0
THETA <- 1

# Sample sizes for each dataset
Y.n <- 50
Z.n <- 50

# Sample standard deviation
Y.sd <- 2
Z.sd <- 1
```

```{r}
# Density function
x <- seq(-8, 8, 0.01)
Y.density <- dnorm(x, PHI, Y.sd)
Z.density <- dnorm(x, PHI + THETA, Z.sd)

# Generate data
set.seed(123)
Y.data <- rnorm(Y.n, PHI, Y.sd)
Z.data <- rnorm(Z.n, PHI + THETA, Z.sd)

# Plot the density functions
df_chart_density <- rbind(
  data.frame(x = x, density = Y.density, dataset = "Y"),
  data.frame(x = x, density = Z.density, dataset = "Z")
)

df_chart_points <- rbind(
  data.frame(x = Y.data, dataset = "Y"),
  data.frame(x = Z.data, dataset = "Z")
)

ggplot(df_chart_density, aes(x, density, color = dataset)) +
  geom_line() +
  geom_point(
    data = df_chart_points,
    aes(x, y = 0, color = dataset),
    alpha = 0.3
  ) +
  theme(aspect.ratio = 1)
```

Generate samples from the first posterior
```{r}
PHI.post.mean <- mean(Y.data)
PHI.post.sd <- Y.sd / sqrt(Y.n)

dpost_phi <- function(x) dnorm(x, PHI.post.mean, PHI.post.sd)
rpost_phi <- function(n) rnorm(n, PHI.post.mean, PHI.post.sd)
```

Conditioning on the posterior mean, generate samples from the conditional posterior of $\theta$
```{r}
dpost_theta_phi <- function(x, phi, log = FALSE) {
  post.mean <- mean(Z.data) - phi
  post.sd <- Z.sd / sqrt(Z.n)
  
  return(dnorm(x, post.mean, post.sd, log = log))
}

rpost_theta_phi <- function(n, phi) {
  post.mean <- mean(Z.data) - phi
  post.sd <- Z.sd / sqrt(Z.n)
  
  return(rnorm(n, post.mean, post.sd))
}
```

Choose 3 values of $\varphi$: closest to the posterior mean, moderate distance from the posterior mean, and very far from the posterior mean.
```{r}
PHI.post.samples <- rpost_phi(1000)
phi.star <- mean(PHI.post.samples)

# Samples from the conditional posterior of theta given phi.star
theta.phi_star <- rpost_theta_phi(1000, phi.star)

# Closest to the mean
idx <- which.min(abs(PHI.post.samples - phi.star))
phi.min <- PHI.post.samples[idx]

# Medium
tmp <- PHI.post.samples[PHI.post.samples > phi.star]
phi.mid <- median(tmp)

# Maximum distance
phi.max <- max(tmp)
```

## Self-Normalised Importance Sampling
```{r}
#' Compute the log ratio between the target and the proposal distributions
compute_log_ratios <- function(samples, log_target_fn, log_proposal_fn) {
  log_target_fn(samples) - log_proposal_fn(samples)
}

#' Compute Self-Normalized Importance Sampling Weights
#' Uses the log-sum-exp trick for numerical stability.
compute_snis_weights <- function(samples, log_target_fn, log_proposal_fn) {
  
  # 1. Compute unnormalized log weights
  log_weights <- compute_log_ratios(samples, log_target_fn, log_proposal_fn)
  
  # 2. Log-Sum-Exp trick to prevent numerical overflow/underflow
  max_log_w <- max(log_weights)
  log_denom <- max_log_w + log(sum(exp(log_weights - max_log_w)))
  
  # 3. Return normalized weights
  exp(log_weights - log_denom)
}

#' Multinomial Resampling
#' Resamples the original samples based on importance weights.
multinomial_resample <- function(samples, weights, num_resamples = NULL) {
  
  # Default to resampling the same number of particles we started with
  n_samples <- if (is.matrix(samples) || is.data.frame(samples)) nrow(samples) else length(samples)
  if (is.null(num_resamples)) num_resamples <- n_samples
  
  # Sample indices with replacement, weighted by our SNIS weights
  resampled_indices <- sample.int(
    n = length(weights), 
    size = num_resamples, 
    replace = TRUE, 
    prob = weights
  )
  
  # Return the subsetted samples
  if (is.matrix(samples) || is.data.frame(samples)) {
    return(samples[resampled_indices, , drop = FALSE])
  } else {
    return(samples[resampled_indices])
  }
}
```

```{r}
# 1. Define your density wrappers
log_target_density   <- function(x) dpost_theta_phi(x, phi.max, log = TRUE)
log_proposal_density <- function(x) dpost_theta_phi(x, phi.star, log = TRUE)

# 1.5 Compute log ratios between target and proposal (for PSIS)
log_ratios <- compute_log_ratios(
  samples = theta.phi_star,
  log_target_fn = log_target_density,
  log_proposal_fn = log_proposal_density
)

# 2. Compute the stable weights
normalized_weights <- compute_snis_weights(
  samples = theta.phi_star, 
  log_target_fn = log_target_density, 
  log_proposal_fn = log_proposal_density
)

# 3. Perform multinomial resampling
resampled_theta <- multinomial_resample(
  samples = theta.phi_star, 
  weights = normalized_weights
)
```

```{r}
# Check PSIS LOO
psis(log_ratios)
```
```{r}
# 1. Prepare theoretical density data
x_grid <- seq(0.25, 1.75, length.out = 500)
df_theory <- data.frame(
  x = x_grid,
  density = dpost_theta_phi(x_grid, phi.min)
)

# 2. Prepare samples data
df_samples <- data.frame(theta = resampled_theta)

# 3. Create the Density Comparison Plot
ggplot() +
  # Empirical SNIS Distribution (KDE)
  geom_density(
    data = df_samples,
    aes(x = theta, fill = "Empirical (SNIS)"),
    alpha = 0.3,
    color = "steelblue"
  ) +
  # Theoretical Target Density
  geom_line(
    data = df_theory,
    aes(x = x, y = density, color = "Theoretical Target")
  ) +
  # Scale and Aesthetic adjustments
  scale_fill_manual(name = "Distribution", values = c("Empirical (SNIS)" = "steelblue")) +
  scale_color_manual(name = "", values = c("Theoretical Target" = "darkred")) +
  theme(legend.position = "bottom", aspect.ratio = 1)
```
Even when PSIS loo is below the prescribed threshold, the SNIS gives a biased estimate of the true distribution.

## Iterative Weighted Moment Matching
When PSIS LOO is above prescribed threshold, we want to trigger IWMM
```{r}
# 1. Define your density wrappers
log_target_density   <- function(draws) dpost_theta_phi(draws, phi.max, log = TRUE)
log_proposal_density <- function(draws) dpost_theta_phi(draws, phi.star, log = TRUE)
log_ratio <- function(draws) log_target_density(draws) - log_proposal_density(draws)

resampled_theta <- moment_match(
  x = as.matrix(theta.phi_star),
  log_prob_prop_fun = log_proposal_density,
  log_ratio_fun = log_ratio,
  k_threshold = 0.7
)
resampled_theta <- resampled_theta$draws[,1]
```

```{r}
# 1. Prepare theoretical density data
x_grid <- seq(0.25, 1.75, length.out = 500)
df_theory <- data.frame(
  x = x_grid,
  density = dpost_theta_phi(x_grid, phi.min)
)

# 2. Prepare samples data
df_samples <- data.frame(theta = resampled_theta)

# 3. Create the Density Comparison Plot
ggplot() +
  # Empirical SNIS Distribution (KDE)
  geom_density(
    data = df_samples,
    aes(x = theta, fill = "Empirical (SNIS)"),
    alpha = 0.3,
    color = "steelblue"
  ) +
  # Theoretical Target Density
  geom_line(
    data = df_theory,
    aes(x = x, y = density, color = "Theoretical Target")
  ) +
  # Scale and Aesthetic adjustments
  scale_fill_manual(name = "Distribution", values = c("Empirical (SNIS)" = "steelblue")) +
  scale_color_manual(name = "", values = c("Theoretical Target" = "darkred")) +
  theme(legend.position = "bottom", aspect.ratio = 1)
```


```{r}
psis_iwmm <- function(samples_s1, rep_val, k_threshold=0.7) {
  S <- length(samples_s1)
  
  log_prop_fn <- function(draws) dpost_theta_phi(draws, phi.star, log = TRUE)
  
  SAMPLES_S2 <- matrix(NA, length(samples_s1), length(samples_s1))
  for (s in 1:S) {
    log_target_fn <- function(draws) dpost_theta_phi(draws, phi.max, log = TRUE)
    log_ratio_fn <- function(draws) log_target_fm(draws) - log_prop_fn(draws)
    
    samples_s2 <- moment_match(
      x = as.matrix(samples_s1),
      log_prob_prop_fun = log_prop_fn,
      log_ratio_fun = log_ratio,
      k_threshold = 0.7
    )
    SAMPLES_S2[s,] <- samples_s2$draws[,1] 
  }
  
  return(SAMPLES_S2)
}

sample_s2 <- as.vector(psis_iwmm(theta.phi_star, phi.star))
```

```{r}
# 1. Prepare theoretical density data
dEP_theta <- function(x) {
  post.mean <- mean(Z.data) - mean(Y.data)
  post.sd <- Y.sd/sqrt(Y.n) + Z.sd/sqrt(Z.n)
  
  dnorm(x, post.mean, post.sd)
}

x_grid <- seq(0.25, 1.75, length.out = 500)
df_theory <- data.frame(
  x = x_grid,
  density = dEP_theta(x_grid)
)

# 2. Prepare samples data
# Note: Apply some thinning
df_samples <- data.frame(theta = sample(sample_s2, size=1000))

# 3. Create the Density Comparison Plot
ggplot() +
  # Empirical SNIS Distribution (KDE)
  geom_density(
    data = df_samples,
    aes(x = theta, fill = "PSIS + IWMM"),
    alpha = 0.3,
    color = "steelblue"
  ) +
  # Theoretical Target Density
  geom_line(
    data = df_theory,
    aes(x = x, y = density, color = "Theoretical Target")
  ) +
  # Scale and Aesthetic adjustments
  scale_fill_manual(name = "Distribution", values = c("PSIS + IWMM" = "steelblue")) +
  scale_color_manual(name = "", values = c("Theoretical Target" = "darkred")) +
  theme(legend.position = "bottom", aspect.ratio = 1)
```



